A Pitman-Yor process prior for estimating shared evolutionary events

### 2021-09-21 manuscript/ms.tex
Update on Overleaf.

\begin{abstract}
    Abstract here \ldots

    \vspace{6pt}
    \noindent\textbf{KEY WORDS:}
\end{abstract}

@@

\section{Introduction}

In this study, we compared the performance of 3 models (Dirichlet prior, Pitman-Yor prior and Uniform prior) 
that are implemented in Ecoevolity software which is the primary software we used to estimate the number of shared divergence time 
among 10 pairs of populations in the past. We simulated data and know the true number of divergence time from it. We are interested 
to estimate the number of divergence time from the generated data. We generated data under 5 various ways to putting prior over all
of the ways that we can assign a pair to divergence event. And then we analyzed the data under 3 models that we can model  
and have some sense of comparison. We are interested in how well the models have estimated the divergence time and how well 
the models have estimated what population pairs belong to which divergence time.

Before going further we needed to assign prior probabilities to 3 models; how probable each model is. Dirichlet-process prior 
assigns prior to a new category with probability of \[\frac{1}{a+1}\] and an exsiting category with probability of \[\frac{a}{a+1}\]. The concentration parameter(a) 
makes the DP flexible because we can adjust the concentration parameter to fit our prior expectation. We can put a distribution on the 
concetration parameter and integrate over uncertainty about the prior probabilities of the divergence models. Pitman-Yor process prior is 
generalization of Dirichlet process. There is a discount parameter(d) added in PYP and when d=0 PYP converges to DP. Dirichlet has an exponential tail 
and the discount parameter gives the PYP flexibity over the tail behavior. Uniform prior is every possible divergence model is equally probable. 
Then we have specified the priors in the config files for each of these 3 models that they are comparable to each other to assess the behavior of the models.

We used these 3 config files and then added 2 more configs (independent and simutanous configs ) to simulate datasets. Independent config 
basically spefifies there are 10 divergence times and each population diverged independently. Simutanous config specifies that the populations 
diverged simutanously and there is 1 shared event. And there is another config file which ignores the constant sites(characters). A total of 6 config 
files are specified to simulate the datasets. We used these 6 config files 4 times to generate scripts and then simulated 10 datasets under each of 
these 5 models with 5000,000 characters. Then we analyzed the data under 3 models and the results show that Pitman-Yor process is doing better job handling 
most difficult scenario all 10 pairs diverging independetly because PYP is more flexible model. The coverage is more robust for Pitman-Yor process twice likely.
So we are interested to see whether this pattern holds with less data. We decided to do another simulation with 4 varying datasets 5,000, 10,000, 50,000, and 100,000. 
All the datasets simulated under 5 models and then analyzed these datasets under 3 models (DP, PYP, and SW). There are equal number of replicates for both simulations 
a total of 720 replicates for each simulation. We found that with varying datasets suggest the same Pitman-Yor process has 2 parameters and more flexible might 
have some advantages especially when it is wrong.

@@

Here we provide a summary of the model implemented in \ecoevolity; please see
\citet{Oaks2018ecoevolity} and \citet{Oaks2019codemog} for a full description.
Our model allows for an arbitrary number of two types of temporal comparisons:
\begin{enumerate}
    \item A population that experienced a change from effective population size
        \epopsize[\rootpopindex]
        to effective size
        \epopsize[\descendantpopindex{}]
        at time \comparisonetime in the past.
        We will refer to this as a \emph{demographic comparison},
        and refer to the population before and after the change in population
        size as ``ancestral'' and ``descendant'', respectively.
    \item A population that diverged at time \comparisonetime in the past into
        two descendant populations, each with unique effective population
        sizes.
        We will refer to this as a \emph{divergence comparison}.
\end{enumerate}
For simplicity, we will use terminology related to divergence comparions, but
all of the theory discussed below also applies to demographic comparisons.

@@

In addition to partition distances as metric for \etimesets error, we used
three additional statistics to quantify performance in estimating \etimesets:
$p(\hat{\etimesets} =\etimesets)$---
    the proportion of simulation replicates in which the true \etimesets had
    the maximum posterior probability,
% However, this metric can be misleading.
% For example, in our simultaneous-divergence and independent-divergence simulations,
% the true divergence model has the maximum prior probability simply because there
% is only a single model with one and 10 divergences.
% Thus, if the data provide no information, there is a 50\% chance of the true
% model having the maximum posterior probability.
$p(\etimesets \given \alldata)$---
    the posterior probability of the true divergence model,
and
$p(\etimesets \in \textrm{95\% CS})$---
    the proportion of simulation replicates for which he true \etimesets is
    included in the 95\% posterior credibility set of divergence models (i.e.,
    coverage).

@@

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the divergence model.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets simulated under the (from left-to-right)
        simultaneous-divergence, DP, PYP, WU, and independent-divergences
        model. 
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which he true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancegrid}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/nchars-model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the independent-divergence
        model as the \dataset size increases.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets with 5,000 to 500,000 characters simulated
        under the independent-divergences model.
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which he true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancebysize}
    \end{center}
\end{figure}

Summary Statistics
Before looking at the results here are some summary statistics that we need to 
understand in order to interpret the results. Coverage or credible sites is the probability
of the number events within the credible sites. The probability of estimated number 
of events equal to the true number of events which basicially tells us how often we 

@@

divergence time ranges from 1 to 10 and 1 they all share the divergence time and 10 they all 
diverged independently. The true number of events is the one has the higest posterior probability.
We have 520 replicates and the grid shows how many of them dall in each cell. The result shows 
that the data geenrated under PYP and analyzed under PYP. There are 88 datasets for which the true 

@@

estimated 4 diverngence times and 1 out of 88 estimated 3 divergence time. 

@@

half of the time there are 10 indepedent divergence times for the datasets that is generated under 
indepdent model. The models are doing very well and they got half of time the correct asnwer. The 

@@

doing little better in terms of how often ge tthe correct answer and also the confident in that 
correct asnwer. The median posterior probability for k=10 getting the correct answer

@@

is an easy scenario and it has the proability of 1 for true number of event versus estimated number of 
event. When the true 1 the models are doing very well estimating the number of divergence time. But 
when the truth is 10 everything is diverging indepdently then it's more difficult scenario for 
modles to handle. Therefore, there is variation in the results. 

@@

which is the correct answer is much lower. But the covarage is unaffacted because the 
likelihood for the missing invariable characters is corrected in Ecoevolity. PYP is doing better when 
dealing with the most difficult scenario all the 10 pairs are diverging independetly as well. Both in
terms of covrage and getting the correct answer is much higher than the other two models.

>>>>

\begin{abstract}
Evolutionary biologists are interested in investigating the evolutionary history of the species. Models are one of the means that allow them to model the world or the evolutionary processes. Therefore, knowing the behavior of the models is important. This study compares the performance of three models Dirichlet process, Pitman-Yor process, and weighted-uniform(DP, PYP, and WU) that are implemented in \ecoevolity software to estimate the number of shared divergence time among 10 pairs of population in the past. Our results show that Pitman-Yor process which is the more flexible model, consistently outperforms the other two models in both simulations (with reduced data sets). The flexibility of Pitman-Yor process has some advantages, especially when it is wrong.

    \vspace{6pt}
    \noindent\textbf{KEY WORDS:Phylogeography, divergence time, evolutionary events, Dirichlet process, Pitman-Yor process, weighted-uniform process, the behavior of models.}

\end{abstract}

@@

\section{Introduction}
Rewrite Intro because it's kind of overlapping with the result section.

Phylogeographic Study \\
Sharing divergence time across taxa \\
Shared divergence across taxa happened via geological events. \\
Temporal clustering reveals evolutionary history of the taxa. \\
Temporal pattern- \\
Model the evolutionary process \\
Important to know the behavior of the models \\
Evolutionary events

We compared the performance of 3 models (Dirichlet prior, Pitman-Yor prior and Uniform prior) 
that are implemented in Ecoevolity software which is the primary software we used to estimate the number of shared divergence time 
among 10 pairs of populations in the past. We simulated data and know the true number of divergence time from it. We are interested 
to estimate the number of divergence time from the generated data. We generated data under 5 various ways to putting prior over all
of the ways that we can assign a pair to divergence event. And then we analyzed the data under 3 models that we can model  
and have some sense of comparison. We are interested in how well the models have estimated the divergence time and how well 
the models have estimated what population pairs belong to which divergence time. We assigned prior probabilities to 3 models; how probable each model is. Dirichlet-process prior 
assigns prior to a new category with probability of \(\dfrac{1}{\concentration +1}\) and an existing category with probability of \(\dfrac{\concentration}{\concentration+1}\). The concentration parameter(\concentration) 
makes the DP flexible because we can adjust the concentration parameter to fit our prior expectation. We can put a distribution on the 
concentration parameter and integrate over uncertainty about the prior probabilities of the divergence models. Pitman-Yor process prior is 
generalization of Dirichlet process. There is a discount parameter($\discount$) added in PYP and when $\discount= 0$ PYP converges to DP. Dirichlet has an exponential tail 
and the discount parameter gives the PYP flexibility over the tail behavior. Uniform prior is every possible divergence model is equally probable. 
Then we have specified the priors in the config files for each of these 3 models that they are comparable to each other to assess the behavior of the models.

We used these 3 config files and then added 2 more configs (independent and simutanous configs) to simulate data sets. Independent config 
basically specifies 10 divergence times and each population diverged independently. Simutanous config specifies that the populations 
diverged simutanously and there is 1 shared event. And there is another config file which ignores the constant sites (characters). A total of 6 config 
files are specified to simulate the data sets. We used these 6 config files 4 times to generate scripts and then simulated 10 data sets under each of 
these 5 models with 5000,000 characters. Then we analyzed the data under 3 models and the results show that Pitman-Yor process is doing better job handling 
most difficult scenario all 10 pairs diverged independently because PYP is more flexible model. The coverage is more robust for Pitman-Yor process twice likely.
So we are interested to see whether this pattern holds with less data. We decided to do another simulation with 4 varying data sets 5,000, 10,000, 50,000, and 100,000. 
All the data sets simulated under 5 models and then analyzed these data sets under 3 models (DP, PYP, and SW). There are equal number of replicates for both simulations 
a total of 720 replicates for each simulation. We found that with varying data sets suggest the same Pitman-Yor process has 2 parameters and more flexible might 
have some advantages especially when it is wrong.

@@

Here we provide a summary of the model implemented in \ecoevolity; please see
\citet{Oaks2018ecoevolity} and \citet{Oaks2019codemog} for a full description.
Our model allows for an arbitrary number of two types of temporal comparisons:
\begin{enumerate}
    \item A population that experienced a change from effective population size
        \epopsize[\rootpopindex]
        to effective size
        \epopsize[\descendantpopindex{}]
        at time \comparisonetime in the past.
        We will refer to this as a \emph{demographic comparison},
        and refer to the population before and after the change in population
        size as ``ancestral'' and ``descendant'', respectively.
    \item A population that diverged at time \comparisonetime in the past into
        two descendant populations, each with unique effective population
        sizes.
        We will refer to this as a \emph{divergence comparison}.
\end{enumerate}
For simplicity, we will use terminology related to divergence comparisons, but
all of the theory discussed below also applies to demographic comparisons.

@@

In addition to partition distances as metric for \etimesets error, we used
three additional statistics to quantify performance in estimating \etimesets:
$p(\hat{\etimesets} =\etimesets)$---
    the proportion of simulation replicates in which the true \etimesets had
    the maximum posterior probability,
% However, this metric can be misleading.
% For example, in our simultaneous-divergence and independent-divergence simulations,
% the true divergence model has the maximum prior probability simply because there
% is only a single model with one and 10 divergences.
% Thus, if the data provide no information, there is a 50\% chance of the true
% model having the maximum posterior probability.
$p(\etimesets \given \alldata)$---
    the posterior probability of the true divergence model,
and
$p(\etimesets \in \textrm{95\% CS})$---
    the proportion of simulation replicates for which the true \etimesets is
    included in the 95\% posterior credibility set of divergence models (i.e.,
    coverage).

@@

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the divergence model.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets simulated under the (from left-to-right)
        simultaneous-divergence, DP, PYP, WU, and independent-divergences
        model. 
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which the true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancegrid}
    \end{center}
\end{figure}

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{../images/from-project-repo/nchars-model-performance-violin-cropped.pdf}
        \captionsetup{listformat=figList}
        \caption{
        Comparing the models' ability to estimate the independent-divergence
        model as the \dataset size increases.
        Each column of plots compares the performance of the Dirichlet process
        (DP), Pitman-Yor process (PYP) and weighted-uniform (WU) models when
        analyzing 720 \datasets with 5,000 to 500,000 characters simulated
        under the independent-divergences model.
        Mean \etimesets error is the mean partition distance
        of the posterior samples from the true
        divergence model (\etimesets),
        $p(\etimesets \given \alldata)$ is the posterior probability
        of the true divergence model,
        $p(\etimesets \given \alldata)$ is the posterior probability of the
        true divergence model,
        and
        $p(\etimesets \in \textrm{95\% CS})$ is the proportion of simulation
        replicates for which the true \etimesets is included in the 95\%
        posterior credibility set of divergence models.
        The bars overlaying the violin plots show the interval between the
        2.5\% and 97.5\% percentiles, and the white dot is the median.
        We generated the plot using matplotlib Version 3.1.3
        \citep{matplotlib}.
        }
        \label{fig:modelperformancebysize}
    \end{center}
\end{figure}

A description of some summary statistics that will help us to interpret the results. Coverage or credible sites is the probability
of the number of events within the credible sites. The probability of estimated number 
of events equal to the true number of events which basically tells us how often we 

@@

divergence time ranges from 1 to 10.  1 refers they all share the divergence time and 10 means they all 
diverged independently. The true number of events is the one has the highest posterior probability.
We have 520 replicates and the grid shows how many of them are in each cell. The result shows 
that the data generated under PYP and analyzed under PYP. There are 88 data sets for which the true 

@@

estimated 4 divergence times and 1 out of 88 estimated 3 divergence time. 

@@

half of the time there are 10 independent divergence times for the data sets that is generated under 
independent model. The models are doing very well and they got half of time the correct answer. The 

@@

doing little better in terms of how often get the correct answer and also the confident in that 
correct answer. The median posterior probability for k=10 getting the correct answer

@@

is an easy scenario and it has the probability of 1 for true number of event versus estimated number of 
event. When the truth is 1 the models are doing very well estimating the number of divergence time. But 
when the truth is 10 everything is diverging independently then it's more difficult scenario for 
models to handle. Therefore, there is variation in the results. 

@@

which is the correct answer is much lower. But the coverage is unaffected because the 
likelihood for the missing invariable characters is corrected in \Ecoevolity. PYP is doing better when 
dealing with the most difficult scenario all the 10 pairs are diverging independently as well. Both in
terms of coverage and getting the correct answer is much higher than the other two models.
